Parsing Postfix log files

### 2007-11-13 doc/logparser.tex
Finished the corrections, or marked where corrections remain to be made. More acronyms, and updated acronyms I missed earlier.

This section provides background information helpful in understanding the
remainder of the paper.  It begins with a discussion of the motivation
underlying the project, followed by some technical information: the use of
a database as an \API\@; a brief introduction to
\ifacronymfirstuse{SMTP}{the}{} \SMTP\@; and a longer introduction to
Postfix, concentrating on the topics most relevant to this paper, namely
Postfix anti-spam restrictions and policy servers.  The assumptions made in
designing and implementing the parser are explained, as are the conventions
used in this paper.  Other projects which attempt to parse Postfix logs are
summarised (full details are available in
appendix~\refwithpage{other-parsers}), finishing with a review of
previously published research in this area.

@@

            \item Triggers can be written to perform actions when certain
                events occur.  In pseudo-SQL\@:

@@

    \item SQL is reasonably standard and many people will already be
        familiar with it; for those unfamiliar with it there are lots of
        readily available resources from which to learn.  Although every
        vendor implements a different dialect of SQL, the basics are the
        same everywhere (analogous to the overall similarities and minor
        differences between British English, American English and
        Australian English).  XXX CITE SOME RESOURCES\@.

@@

The \SMTPlong, originally defined in \RFC{} 821~\cite{RFC821} and updated
in \RFC{} 2821~\cite{RFC2821}, is used for transferring mail between the
sending and receiving \MTA\@.  It is a simple, human readable, plain text
protocol, making it quite easy to test and debug problems with it.  Despite
the simplicity many virus and/or spam sending programs fail to implement it
properly, so requiring strict adherence to the protocol specification is
beneficial in protecting against spam and viruses.\footnote{This violates
the principle of \textit{Be liberal in what you accept, and conservative in
what you send\/} from \RFC{} 760~\cite{rfc760}, but unfortunately that
principle was written in a friendlier time.}  A typical \SMTP{}
conversation resembles the following (the lines starting with a three digit
number are sent by the server, all other lines are sent by the client):

@@

Postfix's design follows the Unix philosophy of \textit{Write programs that
do one thing and do it well\/}~\cite{unix-philosophy}, and is separated
into various component programs to perform the tasks required of an MTA\@:
receive mail, send mail, local delivery of mail, etc. --- full details can
be found at~\cite{postfix-overview}.  Each log line contains the name of
the Postfix program which produced it, and this information is used when
determining which rules should be used to parse each log line (see
section~\refwithpage{rule characteristics} for details.)

@@

Postfix uses several restriction lists (each containing zero or more
restrictions) to decide the fate of an \SMTP{} command, one list for each
stage of the \SMTP{} conversation: client connection, HELO command, MAIL
FROM command, RCPT TO command (possibly multiple times), DATA command, and
end of data.  By default restriction lists will not be evaluated until the
first RCPT TO command is received, because some clients don't deal with
earlier rejections properly; a side benefit of this delay is that Postfix
has more information available when logging the rejection.\footnote{The
\texttt{smtpd\_delay\_reject} parameter in Postfix's configuration controls
this behaviour.}  Postfix uses simple lookup tables as the deciding factor
for some restrictions, e.g.\
\texttt{check\_client\_access~cidr:/etc/postfix/client\_access}
\footnote{\texttt{cidr}~\cite{cidr_table} is the type of the lookup table.}
will check whether the IP address of the connected client matches the left
hand side of each line in the file and will return the right hand side of
the first matching line.  Other restrictions determine their result in
different ways, e.g.\ \texttt{reject\_rbl\_client rbl.example.com} checks
whether the IP address of the client is present in the listed RBL\@.

Each restriction is evaluated to produce a result of \textit{reject},
\textit{permit}, \textit{dunno\/} or the name of another restriction to be
evaluated.\footnote{Other results are possible, for full details
see~\cite{smtpd_access_readme, smtpd_per_user_control, policy-servers}.}
The meaning of \textit{permit\/} and \textit{reject\/} is fairly obvious;
\textit{dunno\/} means to stop evaluating the current restriction and
continue with the next restriction in the list, allowing more specific
cases to be used as exceptions to more general cases.  When the result is
the name of another restriction Postfix will evaluate the new restriction,
allowing restrictions to be chosen based on the client IP address, HELO
hostname, sender address, recipient address, etc.  The administrator can
define new restrictions as a list of existing restrictions, allowing
arbitrarily long and complex sequences of lookups and restrictions.

@@

    \item Greylisting~\cite{greylisting} is a technique that temporarily
        rejects mail when the triple of (sender, recipient, remote IP
        address) is unknown; on second and subsequent delivery attempts
        from that triple the mail will be accepted.  The assumption is that
        maintaining a list of failed addresses and retrying after a
        temporary failure is uneconomical for a spammer, but that a
        legitimate mail server must retry.  Sadly spammers are using
        increasingly complex and well written programs to distribute spam,
        frequently using an \ISP{} provided \SMTP{} server from a
        compromised machine on the \ISP{}'s network.  Greylisting will
        slowly become less useful, but it does block a large percentage of
        spam mail at the moment; the most effective restrictions over the
        93 log files used in testing the parser are shown in
        table~\refwithpage{Summary of rejections}.  Greylisting is
        obviously worth using, at least at the moment, particularly when
        you factor in Greylisting's position as the final restriction which
        a mail must overcome: Greylisting only takes effect for mails which
        have passed every other restriction.

@@

There only appears to be one prior paper published about parsing Postfix
log files: \textit{Log Mail Analyzer: Architecture and Practical
Utilizations\/}~\cite{log-mail-analyser}.  The aim of \LMA{} is quite
different from this parser: it attempts to present correlated log data in a
form suitable for a systems administrator to search using grep(1), awk(1)
and the myriad of other standard Unix text processing utilities already
available, producing a \CSV{} file which can subsequently be imported into
a MySQL or Berkeley DB database.  The decision to support both \CSV{} and
Berkeley DB appears to have been a limiting factor: \CSV{} is a very simple
format where a record is stored in a single line, with fields separated by
a comma or other punctuation symbol.  Problems with CSV files include the
difficulty in escaping separators, providing multiple values for a field
(e.g.\ multiple recipients), adding new fields, and lack of a bundled
schema describing the fields.

Berkeley DB is not an \SQL{} database, supporting only \textbf{(key,
value)} pairs; in the main table the key is an integer referred to by
secondary tables and the value used is a \CSV{} line containing all the
data for that row.  The secondary by-sender, by-recipient, by-date and
by-IP tables use the sender/recipient/date/IP as the key, and a \CSV{} list
of integers referring to the main table is the value.  In effect this
re-implements \SQL{} foreign keys, but without the functionality offered by
even the most basic of \SQL{} databases (joins, ordering, searches, etc.).
It also requires custom code to search on some combination of the above,
though the authors of \LMA{} did provide some queries: IP-STORY,
FROM-STORY, DAILY-EMAIL and DAILY-REJECT\@.  Berkeley DB appears to be the
least useful of the three output formats

The data stored is limited to time and date, hostname and IP address of
client, sender and recipient addresses, and the \SMTP{} code.  Handling of
multiple recipients, \SMTP{} codes or remote servers\footnote{A single mail
may be sent to multiple remote servers if it was addressed to recipients in
different domains, or Postfix needs to try multiple servers for one or more
recipients.} is not explained in the paper, but the options appear to be:

@@

    \item Discard some of the data; this option is obviously undesirable,
        and presumably was not chosen by the LMA authors.

    \item Combine all data into a CSV line, which is then embedded into the
        main CSV line for that mail, making searching even more difficult
        and requiring a second separating character.

@@

The schema used with the MySQL database is undocumented, but at least it's
possible to discover the schema with an SQL database, unlike with Berkeley
DB\@; all \SQL{} databases embed the database schema into the database and
provide commands for displaying it.  Berkeley DB does not embed a schema,
as there is neither requirement nor benefit; it provides \textbf{(key,
value)} pairs, so any additional structuring of the data is imposed by the
application, thus it behoves the application to document this structure.

\parsername{} is bundled with extensive documentation, including an
in-depth explanation of the database schema.  It supports multiple values
for every field for which multiple values are reasonable (e.g.\ the sender
address should have one value only, whereas the recipient address can have
multiple values).  The default database it uses is SQLite3, which supports
most standard SQL features (its lacks full foreign key support), has
libraries available for most common programming languages, and a simple
text based client allowing ad-hoc queries with minimal effort.

@@

% Literature review notes:
%
% Hard-coded parsing, requiring code changes to add more.  Attempts to
% correlate log lines, saves data to database for data mining purposes.
% Hard to extend/expand/understand.  Appears to only save: date and hour,
% DNS name and IP address host, mail server IP address, sender, receiver
% and e-mail status (sent, rejected).  Undocumented schema.  Design
% decision to use CSV as an intermediate format between the log file and
% the database seems to have been restrictive.  Appears to require a
% queueid but majority of log entries (e.g. rejections) lack a queueid.
% Supports whitelisting IP addresses when parsing logs, but whitelisting
% when generating reports/data mining would be preferable.  Supporting
% Berkeley DB is probably limiting the software - an example is the
% difficulty in searching a pipe-delimited string, so they have
% re-implemented foreign keys with tables keyed by ip address etc. pointing
% at the main table - this also won't scale well.  There doesn't appear to
% be any attempt to deal with the complications I've encountered: their
% parsing isn't detailed enough to encounter them.  It doesn't run
% properly; doesn't create any output; throws up errors.

@@

    \item [server\_ip] The IP address (IPv4 or IPv6) of the server: the
        local server when receiving mail, the remote server when sending
        mail.

    \item [server\_hostname] The hostname of the server, it will be
        \texttt{unknown} if the IP address could not be resolved to a
        hostname via DNS\@.

    \item [client\_ip] The client IP address (IPv4 or IPv6): the remote
        server when receiving mail, the local server when sending mail.

    \item [client\_hostname] The hostname of the client, it will be
        \texttt{unknown} if the IP address could not be resolved to a
        hostname via DNS\@.

@@

    \item [data] A field available for anything not covered by other
        fields, e.g.\ the rejection message from an RBL\@.

@@

    \item Sort by regex and visually inspect the list, e.g.\ with SQL
        similar to: \textbf{select regex from rules order by regex;}

@@

    \item [result\_cols, connection\_cols, result\_data and
        connection\_data:] are \newline used by the action to extract and
        save data matched by the regex.

@@

Perl compiles the original \regex{} into an internal representation,
optimising the \regex{} to improve the speed of matching, but this
compilation and optimisation takes CPU time.  Perl automatically caches
static \regexs{}, but dynamic \regexs{} need to be explicitly compiled and
cached.  Graph~\refwithpage{normal regex vs discard regex} shows execution
times with and without caching the regex.  Caching the compiled \regexs{}
is obviously far more efficient; graph~\refwithpage{normal regex vs
discarded regex factor} shows the percentage execution time increase when
not caching each regex.

@@

    \item This is followed by rejection of sender address, recipient
        addresses, client IP address or hostname, etc. (action: rejection;
        flowchart:~4); acceptance of one or more mails (action: clone;
        flowchart:~5); or some interleaving of both.

@@

    \item [CONNECT] Handle a remote client connecting: create a new state
        table entry (indexed by \daemon{smtpd} \pid{}) and save both the
        client hostname and IP address.

@@

If both checks are successful then the mail is assumed to be one of the
offending bogus mails and is discarded.  There will be no further entries
logged for such mails, so without identifying and discarding them they
accumulate in the state table and will cause clashes if the queueid is
reused.  The mail cannot be entered in the database as the only data
available is the client hostname and IP address, but the database schema
requires many more fields be populated (see
sections~\refwithpage{connections table} and~\refwithpage{results table}).

@@

        \url{http://www.thesawmill.co.uk/formats/postfix.html} \newline
        Fields extracted: from, to, server, UID, relay, status, number of
        recipients, origin hostname, origin IP and virus.  The fields
        \textit{server}, \textit{uid\/} and \textit{virus\/} are not
        explained in their documentation: \textit{server\/} is probably the
        server the mail is delivered to, and \textit{uid\/} might be the
        uid of the user submitting mail locally.  Postfix does not perform
        any form of virus checking (though it has many options for
        cooperating with an external virus scanner), so the
        \textit{virus\/} field is a mystery.

        \url{http://www.thesawmill.co.uk/formats/postfix_ii.html} \newline
        Fields extracted: from, to, \RBL{}, client hostname and client
        IP\@.

        \url{http://www.thesawmill.co.uk/formats/beta_postfix.html}
        \newline Fields extracted: from, to, client hostname, client IP,
        relay hostname, relay IP, status, response code, \RBL{} and
        message id.

@@

        AWStats will produce simple graphs for many different services, but
        that restricts it to supporting the Lowest Common Denominator.  The
        data it will extract from an MTA log file is limited to: \newline
        \hspace*{2em} time2, email, email\_r, host, host\_r, method, url,
        code and bytesd.  \newline  There does not seem to be an
        explanation of those fields in the documentation.  AWStats has no
        special purpose code to deal with the intricacies of Postfix logs,
        in fact it operates by coercing Postfix logs into
        Apache\footnote{The Apache web server is the most popular HTTP
        server in use for the past 10 years; more information is available
        at \url{http://httpd.apache.org/}} format log files, for analysis
        by AWStats' HTTP log file parser.  The converting parser is hard
        coded, hard to extend, and only deals with a small portion of the
        log lines generated by Postfix, but it does correlate log lines by
        queueid.

@@

    \item [Anteater] \textit{The Anteater project is a Mail Traffic
        Analyser. Anteater supports currently the logformat produced by
        Sendmail and by Postfix. The tool is written in 100\% C++ and is
        very easy to customize. Input, output, and the analysis are modular
        class objects with a clear interface. There are eight useful
        analyse modules, writing the result in plain ASCII or HTML, to
        stdout or to files.\/}

@@

        YAALA provides an SQL like mini-language used when generating the
        reports; sample reports can be seen at~\cite{yaala-samples}.
        Example query: \newline \hspace*{2em} \texttt{requests BY file
        WHERE host =\~{} Google}. \newline The mini-language is quite
        limited, and cannot be used to extract data for external use,
        merely to create reports.

@@

Graphs are an excellent means of displaying data, transforming a
meaningless stream of numbers into an easily comprehensible form, where
anomalies and patterns are immediately obvious.  These graphs are used to
illustrate the topics discussed in section~\refwithpage{rule efficiency}.
The graphs in the first section cover parser scalability, demonstrating
that performance scales linearly with input size.  The impact of rule
ordering is shown in section~\refwithpage{rule ordering graphs}, and the
anomalous dips and peaks apparent in some graphs are explained.  The next
group of graphs vividly shows the huge impact that caching compiled
\regexs{} has on parser performance.  This section concludes with a
breakdown of the rule hits accumulated during a single test run of the
parser.

@@

The Y axis in the graph above represents the following:

@@

@@

The mail loop was set up by a user modifying his mail forwarding to:
\newline \hspace*{2em}\texttt{$\backslash$username, username@domain}
\newline This instructs Postfix to deliver the mail to the local user, and
also forward to the remote address; this is generally not a problem except
that the remote address in this case is the user's address, creating an
infinite loop.  To prevent this happening Postfix examines the Delivered-To
header in the mail and if the mail has already been delivered to the
current address it is bounced back to the sender with the error message
\texttt{mail forwarding loop for username@domain}.  Ordinarily this works
well, but unfortunately the user noticed they had not received any mail in
a while, and opted to send a test mail to themselves, causing a loop not
caught by Postfix as described below.

@@

    \item Postfix delivers the bounce notification to both the local
        mailbox and to username@domain, as instructed by the user's
        forwarding instructions.  A \texttt{Delivered-To:~username@domain}
        header is added to the forwarded bounce notification, which now has
        an envelope sender address of username@domain.  Log lines are added
        by \daemon{local}, \daemon{qmgr} (twice), \daemon{cleanup}, and
        finally \daemon{pickup}.

@@

The sequence described above occurs extremely rapidly, as Postfix does not
have to deliver the mail to an external system, so mails are delivered,
bounced and generated as fast as the disks can keep up, resulting in a huge
volume of logs.

The vast majority of log lines when a mail loop occurs are from Postfix
components which have a small number of rules associated with them, whereas
in general the most common log lines are logged by \daemon{smtpd}, which
also has the highest number of rules.  \daemon{smtpd} log lines are
distributed across rules much more evenly that \daemon{qmgr},
\daemon{local}, \daemon{cleanup} or \daemon{pickup}, so the average number
of rules required to parse a \daemon{smtpd} log line is much higher that
the average number required to parse other log lines.  Thus the average
number of rules required to parse a random line from logs containing a mail
loop is much lower than the average number required to parse a random log
line from logs without a mail loop, as can be seen in the peaks in
graphs~\refwithpage{execution time vs file size vs number lines
factor},~\refwithpage{percentage increase of shuffled over normal}
and~\refwithpage{percentage increase of reversed over normal}.  The
difference is more extreme in the reversed ordering case because the
largest number of rules to be consulted for the commonly occurring log
lines during a mail loop is far less than the largest number of rules that
are consulted for commonly occurring lines outside of a mail loop.  The
number of rules which needs to be consulted when the ordering is shuffled
varies between the optimum and nadir, and the performance varies
proportionally.

@@

The distribution of log lines across rules is much different and the
average number of rules consulted per log line is much lower when there is
a mail loop; this results in far fewer \regex{} compilations per line than
when there isn't a mail loop, and a correspondingly decreased execution
time.  The increases in execution time when not caching \regexs{} for
different combinations of log files are summarised in the table below:

@@

The graphs presented in this section illustrate the topics discussed in
section~\refwithpage{rule efficiency}.  The first collection of graphs are
about parser scalability, showing the linearly relationship between
execution time and input size.  Section~\refwithpage{rule ordering graphs}
relates to the effect of rule ordering on execution time, and the
unexpected consequences of specific inputs.  The necessity of caching
compiled \regexs{} is attested to by the third group of graphs, where the
difference between caching and discarding is staggering.  This section
concluded with a breakdown of the rule hits accumulated during a single
test run of the parser.

>>>>

This section provides background information helpful in understanding the
remainder of the paper.  It begins with a discussion of the motivation
underlying the project, followed by some technical information: the use of
a database as an \API{}\@; a brief introduction to
\ifacronymfirstuse{SMTP}{the}{} \SMTP{}\@; and a longer introduction to
Postfix, concentrating on the topics most relevant to this paper, namely
Postfix anti-spam restrictions and policy servers.  The assumptions made in
designing and implementing the parser are explained, as are the conventions
used in this paper.  Other projects which attempt to parse Postfix logs are
summarised (full details are available in
appendix~\refwithpage{other-parsers}), finishing with a review of
previously published research in this area.

@@

            \item Triggers can be written to perform actions when certain
                events occur.  In pseudo-\SQL{}\@:

@@

    \item \SQL{} is reasonably standard and many people will already be
        familiar with it; for those unfamiliar with it there are lots of
        readily available resources from which to learn.  Although every
        vendor implements a different dialect of \SQL{}, the basics are the
        same everywhere (analogous to the overall similarities and minor
        differences between British English, American English and
        Australian English).  XXX CITE SOME RESOURCES\@.

@@

The \SMTPlong, originally defined in \RFC{} 821~\cite{RFC821} and updated
in \RFC{} 2821~\cite{RFC2821}, is used for transferring mail between the
sending and receiving \MTA{}\@.  It is a simple, human readable, plain text
protocol, making it quite easy to test and debug problems with it.  Despite
the simplicity many virus and/or spam sending programs fail to implement it
properly, so requiring strict adherence to the protocol specification is
beneficial in protecting against spam and viruses.\footnote{This violates
the principle of \textit{Be liberal in what you accept, and conservative in
what you send\/} from \RFC{} 760~\cite{rfc760}, but unfortunately that
principle was written in a friendlier time.}  A typical \SMTP{}
conversation resembles the following (the lines starting with a three digit
number are sent by the server, all other lines are sent by the client):

@@

Postfix's design follows the Unix philosophy of \textit{Write programs that
do one thing and do it well\/}~\cite{unix-philosophy}, and is separated
into various component programs to perform the tasks required of an
\MTA{}\@: receive mail, send mail, local delivery of mail, etc. --- full
details can be found at~\cite{postfix-overview}.  Each log line contains
the name of the Postfix program which produced it, and this information is
used when determining which rules should be used to parse each log line
(see section~\refwithpage{rule characteristics} for details.)

@@

Postfix uses several restriction lists (each containing zero or more
restrictions) to decide the fate of an \SMTP{} command, one list for each
stage of the \SMTP{} conversation: client connection, HELO command, MAIL
FROM command, RCPT TO command (possibly multiple times), DATA command, and
end of data.  By default restriction lists will not be evaluated until the
first RCPT TO command is received, because some clients don't deal with
earlier rejections properly; a side benefit of this delay is that Postfix
has more information available when logging the rejection.\footnote{The
\texttt{smtpd\_delay\_reject} parameter in Postfix's configuration controls
this behaviour.}  Postfix uses simple lookup tables as the deciding factor
for some restrictions, e.g.\
\texttt{check\_client\_access~cidr:/etc/postfix/client\_access}
\footnote{\texttt{cidr}~\cite{cidr_table} is the type of the lookup table.}
will check whether the \IP{} address of the connected client matches the
left hand side of each line in the file and will return the right hand side
of the first matching line.  Other restrictions determine their result in
different ways, e.g.\ \texttt{reject\_rbl\_client rbl.example.com} checks
whether the \IP{} address of the client is present in the listed \RBL{}\@.

Each restriction is evaluated to produce a result of \textit{reject},
\textit{permit}, \textit{dunno\/} or the name of another restriction to be
evaluated.\footnote{Other results are possible, for full details
see~\cite{smtpd_access_readme, smtpd_per_user_control, policy-servers}.}
The meaning of \textit{permit\/} and \textit{reject\/} is fairly obvious;
\textit{dunno\/} means to stop evaluating the current restriction and
continue with the next restriction in the list, allowing more specific
cases to be used as exceptions to more general cases.  When the result is
the name of another restriction Postfix will evaluate the new restriction,
allowing restrictions to be chosen based on the client \IP{} address, HELO
hostname, sender address, recipient address, etc.  The administrator can
define new restrictions as a list of existing restrictions, allowing
arbitrarily long and complex sequences of lookups and restrictions.

@@

    \item Greylisting~\cite{greylisting} is a technique that temporarily
        rejects mail when the triple of (sender, recipient, remote \IP{}
        address) is unknown; on second and subsequent delivery attempts
        from that triple the mail will be accepted.  The assumption is that
        maintaining a list of failed addresses and retrying after a
        temporary failure is uneconomical for a spammer, but that a
        legitimate mail server must retry.  Sadly spammers are using
        increasingly complex and well written programs to distribute spam,
        frequently using an \ISP{} provided \SMTP{} server from a
        compromised machine on the \ISP{}'s network.  Greylisting will
        slowly become less useful, but it does block a large percentage of
        spam mail at the moment; the most effective restrictions over the
        93 log files used in testing the parser are shown in
        table~\refwithpage{Summary of rejections}.  Greylisting is
        obviously worth using, at least at the moment, particularly when
        you factor in Greylisting's position as the final restriction which
        a mail must overcome: Greylisting only takes effect for mails which
        have passed every other restriction.

@@

There only appears to be one prior paper published about parsing Postfix
log files: \textit{Log Mail Analyzer: Architecture and Practical
Utilizations\/}~\cite{log-mail-analyser}.  The aim of \LMA{} is quite
different from this parser: it attempts to present correlated log data in
a form suitable for a systems administrator to search using grep(1), awk(1)
and the myriad of other standard Unix text processing utilities already
available, producing a \CSV{} file which can subsequently be imported into
a MySQL or Berkeley DB database.  The decision to support both \CSV{} and
Berkeley DB appears to have been a limiting factor: \CSV{} is a very simple
format where a record is stored in a single line, with fields separated by
a comma or other punctuation symbol.  Problems with \CSV{} files include
the difficulty in escaping separators, providing multiple values for a
field (e.g.\ multiple recipients), adding new fields, and lack of a bundled
schema describing the fields.

Berkeley DB is not an \SQL{} database, supporting only \textbf{(key,
value)} pairs; in the main table the key is an integer referred to by
secondary tables and the value used is a \CSV{} line containing all the
data for that row.  The secondary by-sender, by-recipient, by-date and
by-\IP{} tables use the sender/recipient/date/\IP{} as the key, and a
\CSV{} list of integers referring to the main table is the value.  In
effect this re-implements \SQL{} foreign keys, but without the
functionality offered by even the most basic of \SQL{} databases (joins,
ordering, searches, etc.).  It also requires custom code to search on some
combination of the above, though the authors of \LMA{} did provide some
queries: IP-STORY, FROM-STORY, DAILY-EMAIL and DAILY-REJECT\@.  Berkeley DB
appears to be the least useful of the three output formats

The data stored is limited to time and date, hostname and \IP{} address of
client, sender and recipient addresses, and the \SMTP{} code.  Handling of
multiple recipients, \SMTP{} codes or remote servers\footnote{A single mail
may be sent to multiple remote servers if it was addressed to recipients in
different domains, or Postfix needs to try multiple servers for one or more
recipients.} is not explained in the paper, but the options appear to be:

@@

    \item Discard some of the data; this option is obviously undesirable,
        and presumably was not chosen by the \LMA{} authors.

    \item Combine all data into a \CSV{} line, which is then embedded into
        the main \CSV{} line for that mail, making searching even more
        difficult and requiring a second separating character.

@@

The schema used with the MySQL database is undocumented, but at least it's
possible to discover the schema with an \SQL{} database, unlike with
Berkeley DB\@; all \SQL{} databases embed the database schema into the
database and provide commands for displaying it.  Berkeley DB does not
embed a schema, as there is neither requirement nor benefit; it provides
\textbf{(key, value)} pairs, so any additional structuring of the data is
imposed by the application, thus it behoves the application to document
this structure.

\parsername{} is bundled with extensive documentation, including an
in-depth explanation of the database schema.  It supports multiple values
for every field for which multiple values are reasonable (e.g.\ the sender
address should have one value only, whereas the recipient address can have
multiple values).  The default database it uses is SQLite3, which supports
most standard \SQL{} features (its lacks full foreign key support), has
libraries available for most common programming languages, and a simple
text based client allowing ad-hoc queries with minimal effort.

@@

% Literature review notes:
%
% Hard-coded parsing, requiring code changes to add more.  Attempts to
% correlate log lines, saves data to database for data mining purposes.
% Hard to extend/expand/understand.  Appears to only save: date and hour,
% \DNS{} name and \IP{} address host, mail server \IP{} address, sender,
% receiver and e-mail status (sent, rejected).  Undocumented schema.
% Design decision to use \CSV{} as an intermediate format between the log
% file and the database seems to have been restrictive.  Appears to require
% a queueid but majority of log entries (e.g. rejections) lack a queueid.
% Supports whitelisting \IP{} addresses when parsing logs, but whitelisting
% when generating reports/data mining would be preferable.  Supporting
% Berkeley DB is probably limiting the software - an example is the
% difficulty in searching a pipe-delimited string, so they have
% re-implemented foreign keys with tables keyed by ip address etc. pointing
% at the main table - this also won't scale well.  There doesn't appear to
% be any attempt to deal with the complications I've encountered: their
% parsing isn't detailed enough to encounter them.  It doesn't run
% properly; doesn't create any output; throws up errors.

@@

    \item [server\_ip] The \IP{} address (IPv4 or IPv6) of the server: the
        local server when receiving mail, the remote server when sending
        mail.

    \item [server\_hostname] The hostname of the server, it will be
        \texttt{unknown} if the \IP{} address could not be resolved to a
        hostname via \DNS{}\@.

    \item [client\_ip] The client \IP{} address (IPv4 or IPv6): the remote
        server when receiving mail, the local server when sending mail.

    \item [client\_hostname] The hostname of the client, it will be
        \texttt{unknown} if the \IP{} address could not be resolved to a
        hostname via \DNS{}\@.

@@

    \item [data] A field available for anything not covered by other
        fields, e.g.\ the rejection message from an \RBL{}\@.

@@

    \item Sort by \regex{} and visually inspect the list, e.g.\ with \SQL{}
        similar to: \textbf{select regex from rules order by regex;}

@@

    \item [result\_cols, connection\_cols, result\_data and
        connection\_data:] are \newline used by the action to extract and
        save data matched by the \regex{}.

@@

Perl compiles the original \regex{} into an internal representation,
optimising the \regex{} to improve the speed of matching, but this
compilation and optimisation takes CPU time.  Perl automatically caches
static \regexs{}, but dynamic \regexs{} need to be explicitly compiled and
cached.  Graph~\refwithpage{normal regex vs discard regex} shows execution
times with and without caching the \regex{}.  Caching the compiled
\regexs{} is obviously far more efficient; graph~\refwithpage{normal regex
vs discarded regex factor} shows the percentage execution time increase
when not caching each \regex{}.

@@

    \item This is followed by rejection of sender address, recipient
        addresses, client \IP{} address or hostname, etc. (action:
        rejection; flowchart:~4); acceptance of one or more mails (action:
        clone; flowchart:~5); or some interleaving of both.

@@

    \item [CONNECT] Handle a remote client connecting: create a new state
        table entry (indexed by \daemon{smtpd} \pid{}) and save both the
        client hostname and \IP{} address.

@@

If both checks are successful then the mail is assumed to be one of the
offending bogus mails and is discarded.  There will be no further entries
logged for such mails, so without identifying and discarding them they
accumulate in the state table and will cause clashes if the queueid is
reused.  The mail cannot be entered in the database as the only data
available is the client hostname and \IP{} address, but the database schema
requires many more fields be populated (see
sections~\refwithpage{connections table} and~\refwithpage{results table}).

@@

        \url{http://www.thesawmill.co.uk/formats/postfix.html} \newline
        Fields extracted: from, to, server, UID, relay, status, number of
        recipients, origin hostname, origin \IP{} and virus.  The fields
        \textit{server}, \textit{uid\/} and \textit{virus\/} are not
        explained in their documentation: \textit{server\/} is probably the
        server the mail is delivered to, and \textit{uid\/} might be the
        uid of the user submitting mail locally.  Postfix does not perform
        any form of virus checking (though it has many options for
        cooperating with an external virus scanner), so the
        \textit{virus\/} field is a mystery.

        \url{http://www.thesawmill.co.uk/formats/postfix_ii.html} \newline
        Fields extracted: from, to, \RBL{}, client hostname and client
        \IP{}\@.

        \url{http://www.thesawmill.co.uk/formats/beta_postfix.html}
        \newline Fields extracted: from, to, client hostname, client \IP{},
        relay hostname, relay \IP{}, status, response code, \RBL{} and
        message id.

@@

        AWStats will produce simple graphs for many different services, but
        that restricts it to supporting the Lowest Common Denominator.  The
        data it will extract from an \MTA{} log file is limited to:
        \newline \hspace*{2em} time2, email, email\_r, host, host\_r,
        method, url, code and bytesd.  \newline  There does not seem to be
        an explanation of those fields in the documentation.  AWStats has
        no special purpose code to deal with the intricacies of Postfix
        logs, in fact it operates by coercing Postfix logs into
        Apache\footnote{The Apache web server is the most popular HTTP
        server in use for the past 10 years; more information is available
        at \url{http://httpd.apache.org/}} format log files, for analysis
        by AWStats' HTTP log file parser.  The converting parser is hard
        coded, hard to extend, and only deals with a small portion of the
        log lines generated by Postfix, but it does correlate log lines by
        queueid.

@@

    \item [Anteater] \textit{The Anteater project is a Mail Traffic
        Analyser. Anteater supports currently the logformat produced by
        Sendmail and by Postfix. The tool is written in 100\% C++ and is
        very easy to customize. Input, output, and the analysis are modular
        class objects with a clear interface. There are eight useful
        analyse modules, writing the result in plain ASCII or \HTML{}, to
        stdout or to files.\/}

@@

        YAALA provides an \SQL{} like mini-language used when generating
        the reports; sample reports can be seen at~\cite{yaala-samples}.
        Example query: \newline \hspace*{2em} \texttt{requests BY file
        WHERE host =\~{} Google}. \newline The mini-language is quite
        limited, and cannot be used to extract data for external use,
        merely to create reports.

@@

Graphs are an excellent means of displaying data, transforming a
meaningless stream of numbers into an easily comprehensible form, where
anomalies and patterns are immediately obvious.  These graphs are used to
illustrate the topics discussed in section~\refwithpage{rule efficiency}.
The graphs in the first section cover parser scalability, demonstrating
that performance scales linearly with input size.  The impact of rule
ordering is shown in section~\refwithpage{rule ordering graphs}, and the
anomalous dips and peaks apparent in some graphs are explained.  The third
group of graphs vividly shows the huge impact that caching compiled
\regexs{} has on parser performance.  This section concludes with a
breakdown of the rule hits accumulated during a single test run of the
parser.

@@

The Y axis in graph~\refwithpage{execution time vs file size vs number of
lines graph} represents the following:

@@

Graph~\refwithpage{execution time vs file size vs number lines factor}
shows the ratio of execution time vs file size and number of lines (higher
is better, it means more bytes or lines processed per second).  The ratios
are quite tightly banded with the exception of log files 22 and 62--68,
where they are noticeably higher; graph~\refwithpage{execution time vs file
size vs number lines factor} and table~\refwithpage{execution time vs file
size vs number lines factor table} show that the parser's execution time
scales linearly with input size.

@@

The mail loop was set up by a user modifying his mail forwarding to:
\newline \hspace*{2em}\texttt{$\backslash$username, username@domain}
\newline This instructs Postfix to deliver the mail to the local user, and
also forward it to the remote address; this is generally not a problem
except that the remote address in this case is the user's address, creating
an infinite loop.  To prevent this happening Postfix examines the
Delivered-To header in the mail, and if the mail has already been delivered
to the current address it is bounced back to the sender with the error
message \texttt{mail forwarding loop for username@domain}.  Ordinarily this
works well, but unfortunately in this case the user noticed they had not
received any mail in a while and opted to send a test mail to themselves,
causing a loop not caught by Postfix as described below.

@@

    \item Postfix accepts the bounce notification and delivers it to both
        the local mailbox and to username@domain, as instructed by the
        user's forwarding instructions.  A
        \texttt{Delivered-To:~username@domain} header is added to the
        forwarded bounce notification, which now has an envelope sender
        address of username@domain.  Log lines are added by \daemon{local},
        \daemon{qmgr} (twice), \daemon{cleanup}, and finally
        \daemon{pickup}.

@@

The sequence described above occurs extremely rapidly because Postfix does
not have to deliver the mail to an external system, so mails are delivered,
bounced and generated as fast as the disks can keep up, resulting in a huge
volume of logs.

The vast majority of log lines when a mail loop occurs are from Postfix
components which have a small number of rules associated with them, whereas
in general \daemon{smtpd} adds the majority of log lines, and also has the
highest number of rules.  \daemon{smtpd} log lines are distributed across
rules much more evenly than the log lines of \daemon{qmgr}, \daemon{local},
\daemon{cleanup} or \daemon{pickup}, so the average number of rules
required to parse a \daemon{smtpd} log line is much higher that the average
number required to parse other log lines.

These two characteristics combine to reduce the average number or rules
required to parse a log line when there is a mail loop, as shown by the
peaks in graph~\refwithpage{execution time vs file size vs number lines
factor}.  When the rule ordering is reversed the majority of log lines
generated by a mail loop will be parsed with very few rules, whereas
without a mail loop the majority of log lines require a large number of
rules; this leads to a noticeable drop in the average time required to
parse a log line, as shown in graph~\refwithpage{percentage increase of
reversed over normal}.  The number of rules which need to be consulted when
the ordering is shuffled varies between the optimum and nadir, and the
performance varies proportionally.

@@

The distribution of log lines across rules when there is a mail loop is
much different and the average number of rules consulted per log line is
much lower; this results in far fewer \regex{} compilations per line than
when there isn't a mail loop, and a correspondingly decreased execution
time.  The increases in execution time when not caching \regexs{} for
different combinations of log files are summarised in the table below:

@@

The graphs presented in this section illustrate the topics discussed in
section~\refwithpage{rule efficiency}.  The first collection of graphs are
about parser scalability, showing the linear relationship between execution
time and input size.  Section~\refwithpage{rule ordering graphs} relates to
the effect of rule ordering on execution time, and the unexpected
consequences of specific inputs.  The necessity of caching compiled
\regexs{} is attested to by the third group of graphs, where the difference
between caching and discarding is staggering.  This section concluded with
a breakdown of the rule hits accumulated during a single test run of the
parser.
